Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-17 vertical-thermal-review.Rmd
1st abbreviation

A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and crown position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019]. 
This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature might also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles. 
Vertical transport at the macroscopic scale of a forest canopy is dominated by turbulent transport rather than by molecular diffusion. 
Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) that depend on both friction and buoyancy [@penmanWeatherWheatEssay1960]. 
In this “K-theory” regime, wind speed declines logarithmically with proximity to the ground or canopy surface [@monteithPrinciplesEnvironmentalPhysics2013]. Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
However, K-theory often fails within canopies, making forest verticcal wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991]. 
These "sweep-eject" events can couple understory conditions more directly to conditions above the canopy than beneath.

>>>>

A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and crown position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019]. 
This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature might also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles. 
Vertical transport at the macroscopic scale of a forest canopy is dominated by turbulent transport rather than by molecular diffusion. 
Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) that depend on both friction and buoyancy [@penmanWeatherWheatEssay1960]. 
In this “K-theory” regime, wind speed declines logarithmically with proximity to the ground or canopy surface [@monteithPrinciplesEnvironmentalPhysics2013]. Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
However, K-theory often fails within canopies, making forest vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991]. 
These "sweep-eject" events can couple understory conditions more directly to conditions above the canopy than beneath.
